---
title: "From segments to a density surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From segments to a density surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(distsamp)
```

`distsamp` stops at segments. This vignette picks up where
`vignette("segmenting-narwc-data")` ends and shows how to hand its output to
`Distance` and `dsm`, because the handoff is where the units and the
segment-to-detection keying are easy to get wrong.

The `Distance` and `dsm` chunks below are **not evaluated** — neither package is
a dependency of `distsamp`, so this vignette builds without them. The code is
real and runnable once you install them.

**The data is mock.** `narwc-example.csv` is a synthetic survey generated by
`data-raw/make-fixture.R`; no real NARWC records are distributed with this
package or were used to develop it. It has two days, seven tracks, and a handful
of sightings — nowhere near enough to fit a detection function or a density
surface to. Every number below illustrates the plumbing, and none of it is a
result.

## What you start with

```{r}
path <- system.file("extdata", "narwc-example.csv", package = "distsamp")
dat <- read_narwc(path)
segs <- segment_survey(dat, seg_length = 5, seed = 1, species = "RIWH")
segs
```

Two tables matter here.

**`segments`** — the sampling units. One row each, with effort and a location:

```{r}
head(segs$segments[, c("seg_id", "seg_eff", "mid_lat", "mid_lon", "wt_beaufort")])
```

**`detections`** — one row per sighting, with perpendicular distance:

```{r}
segs$detections[, c("seg_id", "SPECCODE", "size", "distance", "side")]
```

`seg_id` is the key between them. Every detection belongs to exactly one
segment, and every segment carries the effort that detection was made on.

## The units trap

This is the single most common way to get a density estimate wrong by three
orders of magnitude, so it is worth being explicit.

```{r}
c(effort = "km", distance = segs$settings$distance_units)
```

`seg_eff` is in **kilometres**, because that is what effort accumulation
produces. `distance` defaults to **metres**, because `ALT` is in metres
(handbook 8.A.1) and perpendicular distances for aerial cetacean surveys are
conventionally reported that way.

You have two options. Either work in kilometres throughout:

```{r}
segs_km <- segment_survey(dat, seg_length = 5, seed = 1, species = "RIWH",
                          distance_units = "km")
segs_km$detections$distance
```

or keep metres and tell `Distance` about it via `convert_units`. For effort in
km, distance in m, and area in km², that conversion factor is:

```{r}
# (m -> km) * 2 sides * (km effort)
convert_units <- 0.001
```

Whichever you pick, state it in your methods.

## Look at the distances first

Before fitting anything, look at the distribution you are about to fit a curve
to.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
plot(segs, what = "distances")
```

You are looking for a shoulder near zero and a tail that falls away. What you
find changes what you fit:

- **A dip in the first bin** may not be a sampling artefact. An aircraft cannot
  see the water directly beneath it — for the Skymaster, handbook 8.A.31 says
  CETAP-era distances "are actually measured from about 1/8 mile to either side
  of the survey line". That argues for `ds(left = ...)`, not for assuming the
  animals were not there.
- **A peak away from zero** is what the gamma key function describes. It is not
  available in `Distance::ds()`, which offers `hn`, `hr`, and `unif` only; it
  lives in `mrds::ddf(key = "gamma")`. Note that a gamma key and a left
  truncation model the same phenomenon, so using both counts the blind spot
  twice.
- **A long flat tail** is the case for truncating harder.
- **A distance the aircraft could not possibly have seen** is not a wide
  detection, it is a broken one — a sighting matched to the wrong segment, or a
  coordinate with a lost sign. The axis stops at the 99th percentile so one such
  value cannot push every real distance into the first bin, and the subtitle
  reports how many lie beyond and how large the largest is. If it says the
  largest is too far to be a detection, fix that before fitting anything:
  `?sighting_distances` says where distances come from and which source produced
  it.

None of this tells you anything about `g(0)`. Animals that were submerged when
the aircraft passed leave no trace in this plot, and a perfectly smooth curve is
equally consistent with half of them having been invisible.

## Fitting a detection function

`Distance::ds()` wants a data frame with a `distance` column, optionally with
`size` and detection covariates. `segs$detections` is already that shape.

`detection_data()` assembles that table for you:

```{r}
flat <- detection_data(
  segs,
  area = 5811,                    # square km - yours to supply
  covariates = "wt_beaufort"
)
head(flat)
```

```{r, eval = FALSE}
library(Distance)

df <- ds(flat, key = "hn", truncation = "10%", formula = ~ wt_beaufort)

summary(df)
plot(df)
gof_ds(df)
```

Four things worth pausing on.

**`area` is required and has no default.** It scales abundance directly. The
scripts this package was rewritten from had 5,811 km² written into the middle of
a function, which is exactly how a study-area figure outlives the study it
belonged to.

**Segments with no detections are in the table, carrying a missing `distance`.**

```{r}
sum(is.na(flat$distance))
```

That is not padding. It is how a flatfile records effort that produced no
sightings, and dropping those rows would remove the denominator and inflate
density.

**Use `wt_beaufort`, not `mean_beaufort`.** The weighted version weights each
record by the distance it covers, so a sea state held over 3 km counts for more
than one recorded over 200 m. That is the quantity detection actually depends on.

**Circling detections are excluded by default**, and the exclusion is reported
rather than silent. A position logged off the track is not measured at the moment
of detection. They still count towards the segment's abundance, so this does not
touch `segs$sightings` — excluding a detection from the detection function while
counting the animal towards density is a normal combination, and the two are
separate arguments:

```{r}
nrow(detection_data(segs, area = 5811, include_circling = TRUE))
```

**Check which source each distance came from.** A methods section has to state
this, and point and interval distances cannot be fitted together:

```{r}
table(flat$distance_source, useNA = "no")
```

**Truncation discards the tail, but not the effort.** `detection_data()` takes a
`truncation` argument that drops detections beyond it and leaves their segments
in place with effort intact — effort searched is effort searched whether or not
anything was seen within the truncation distance. Pass the same value to `ds()`,
or you will fit the model to one set of detections and scale abundance by
another.

```{r}
nrow(detection_data(segs, area = 5811, truncation = 500))
```

### Comparing several models

Fitting a set and picking the best is standard, and there are two traps in it.

**AIC only compares models fitted to the same data.** Changing the truncation
changes which detections are in the likelihood, so AICs either side of that are
not comparable. Vary key function, adjustments, and covariates within a fixed
truncation; treat truncation as a separate comparison altogether.

**Binned and unbinned fits cannot share a table either.** `STRIP`-derived
distances are intervals and are fitted binned; angle- and position-derived
distances are points. A multi-year model set splits along that line whether you
want it to or not.

And rank on more than AIC. What propagates into abundance is the effective strip
half-width, and two models within 2 AIC of each other can give materially
different values of it — so read $\bar{p}$ and its CV alongside the AIC, with a
goodness-of-fit statistic.

The package repository's `docs/07-fitting-architecture.md` works through this,
along with where an automated model-selection sweep should live and what can and
cannot be done about `g(0)`.

## Building the `dsm` inputs

`dsm` wants two data frames with specific column names. The mapping from
`distsamp` output is mechanical:

```{r, eval = FALSE}
library(dsm)

segment.data <- data.frame(
  Sample.Label = segs$segments$seg_id,
  Effort       = segs$segments$seg_eff,   # km
  X            = segs$segments$mid_lon,   # project these first, see below
  Y            = segs$segments$mid_lat,
  beaufort     = segs$segments$wt_beaufort
)

observation.data <- data.frame(
  object       = seq_len(nrow(df_data)),
  Sample.Label = df_data$seg_id,
  size         = df_data$size,
  distance     = df_data$distance
)
```

`object` is just a detection identifier; `dsm` uses it to link a detection to
its row in the fitted detection function, so it must match the row order of the
data you passed to `ds()`.

### Project the coordinates first

`mid_lon`/`mid_lat` are decimal degrees. Fitting a smooth of longitude and
latitude in degrees distorts distances badly at the latitudes NARWC surveys
cover — a degree of longitude is about 81 km at 43°N against 111 km for a degree
of latitude. Project to something equal-area before modelling:

```{r, eval = requireNamespace("sf", quietly = TRUE)}
pts <- segments_as_sf(segs, "midpoints")

# Albers equal-area, Gulf of Maine
gom_aea <- paste(
  "+proj=aea +lat_1=40 +lat_2=46 +lat_0=43 +lon_0=-69",
  "+x_0=0 +y_0=0 +datum=WGS84 +units=km +no_defs"
)
projected <- sf::st_transform(pts, gom_aea)

xy <- sf::st_coordinates(projected)
head(xy)
```

Then use those `X`/`Y` in `segment.data`.

## Fitting the model

```{r, eval = FALSE}
mod <- dsm(
  count ~ s(X, Y, bs = "ts") + s(depth, bs = "ts"),
  ddf.obj = df,
  segment.data = segment.data,
  observation.data = observation.data,
  family = tw(),          # Tweedie handles the zero inflation
  method = "REML"
)

summary(mod)
plot(mod)
gam.check(mod)
```

A Tweedie or negative binomial family is usually right: most segments hold zero
animals, and the non-zero ones are overdispersed. Miller et al. (2013) discuss
the choice.

## Covariates are your job

`distsamp` gives you a location per segment and nothing else. Environmental
covariates — depth, sea surface temperature, fronts — have to be sampled at each
segment midpoint yourself. `segments_as_sf(segs, "midpoints")` is the intended
handoff:

```{r, eval = FALSE}
pts <- segments_as_sf(segs, "midpoints")

# e.g. with terra
depth <- terra::extract(bathymetry_raster, terra::vect(pts))
segment.data$depth <- depth[, 2]
```

Two cautions.

Sample at the **midpoint**, which is what Becker et al. (2019) do — covariates
"derived based on the segment's geographical mid-point", with SST and depth
standard deviations taken over a 3 × 3-pixel box around it. `segment_midpoints()`
places that point half the segment's *effort* along the track rather than at the
centroid of its records, so it is on the trackline by construction.

And match covariates in **time** as well as space. A segment flown in April
needs April's SST, not the annual mean. `segs$segments$DATE` carries the survey
date for exactly this.

## Reporting

Record the `seed`. Without it the segmentation is not reproducible, and neither
is anything downstream of it.

```{r}
str(segs$settings)
```

Suggested methods wording is in `docs/06-references.md` in the package
repository.

## References

Becker, E.A., Forney, K.A., Ferguson, M.C., Foley, D.G., Smith, R.C., Barlow, J.
and Redfern, J.V. (2010) Comparing California Current cetacean-habitat models
developed using in situ and remotely sensed sea surface temperature data.
*Marine Ecology Progress Series* 413:163-183.
<https://doi.org/10.3354/meps08696>

Becker, E.A., Forney, K.A., Redfern, J.V., Barlow, J., Jacox, M.G., Roberts, J.J.
and Palacios, D.M. (2019) Predicting cetacean abundance and distribution in a
changing climate. *Diversity and Distributions* 25:626-643.
<https://doi.org/10.1111/ddi.12867>

Hedley, S.L. and Buckland, S.T. (2004) Spatial models for line transect sampling.
*Journal of Agricultural, Biological, and Environmental Statistics* 9:181-199.
<https://doi.org/10.1198/1085711043578>

Miller, D.L., Burt, M.L., Rexstad, E.A. and Thomas, L. (2013) Spatial models for
distance sampling data: recent developments and future directions. *Methods in
Ecology and Evolution* 4:1001-1010.
<https://doi.org/10.1111/2041-210X.12105>

Miller, D.L., Rexstad, E., Thomas, L., Marshall, L. and Laake, J.L. (2019)
Distance sampling in R. *Journal of Statistical Software* 89(1):1-28.
<https://doi.org/10.18637/jss.v089.i01>
