---
title: "From segments to a density surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From segments to a density surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(distsamp)
```

`distsamp` stops at segments. This vignette picks up where
`vignette("segmenting-narwc-data")` ends and shows how to hand its output to
`Distance` and `dsm`, because the handoff is where the units and the
segment-to-detection keying are easy to get wrong.

The `Distance` and `dsm` chunks below are **not evaluated** — neither package is
a dependency of `distsamp`, so this vignette builds without them. The code is
real and runnable once you install them.

## What you start with

```{r}
path <- system.file("extdata", "narwc-example.csv", package = "distsamp")
dat <- read_narwc(path)
segs <- segment_survey(dat, seg_length = 5, seed = 1, species = "RIWH")
segs
```

Two tables matter here.

**`segments`** — the sampling units. One row each, with effort and a location:

```{r}
head(segs$segments[, c("seg_id", "seg_eff", "mid_lat", "mid_lon", "wt_beaufort")])
```

**`detections`** — one row per sighting, with perpendicular distance:

```{r}
segs$detections[, c("seg_id", "SPECCODE", "size", "distance", "side")]
```

`seg_id` is the key between them. Every detection belongs to exactly one
segment, and every segment carries the effort that detection was made on.

## The units trap

This is the single most common way to get a density estimate wrong by three
orders of magnitude, so it is worth being explicit.

```{r}
c(effort = "km", distance = segs$settings$distance_units)
```

`seg_eff` is in **kilometres**, because that is what effort accumulation
produces. `distance` defaults to **metres**, because `ALT` is in metres
(handbook 8.A.1) and perpendicular distances for aerial cetacean surveys are
conventionally reported that way.

You have two options. Either work in kilometres throughout:

```{r}
segs_km <- segment_survey(dat, seg_length = 5, seed = 1, species = "RIWH",
                          distance_units = "km")
segs_km$detections$distance
```

or keep metres and tell `Distance` about it via `convert_units`. For effort in
km, distance in m, and area in km², that conversion factor is:

```{r}
# (m -> km) * 2 sides * (km effort)
convert_units <- 0.001
```

Whichever you pick, state it in your methods.

## Fitting a detection function

`Distance::ds()` wants a data frame with a `distance` column, optionally with
`size` and detection covariates. `segs$detections` is already that shape.

```{r, eval = FALSE}
library(Distance)

df_data <- segs$detections[!is.na(segs$detections$distance), ]

# Attach segment-level detection covariates by seg_id
df_data <- merge(
  df_data,
  segs$segments[, c("seg_id", "wt_beaufort")],
  by = "seg_id"
)

df <- ds(
  df_data,
  key = "hn",
  truncation = "10%",
  formula = ~ wt_beaufort
)

summary(df)
plot(df)
gof_ds(df)
```

Three things worth pausing on.

**Use `wt_beaufort`, not `mean_beaufort`.** The weighted version weights each
record by the distance it covers, so a sea state held over 3 km counts for more
than one recorded over 200 m. That is the quantity detection actually depends on.

**Circling detections have no distance and drop out here.** That is correct —
a position logged off the track is not a perpendicular distance. They still
count towards the segment's abundance, so do not remove them from
`segs$sightings`.

```{r}
subset(segs$detections, is.na(distance))
```

**Truncation discards the tail.** Detections beyond the truncation distance are
excluded from the detection function *and* must be excluded from the counts, or
abundance and effective search area will disagree.

## Building the `dsm` inputs

`dsm` wants two data frames with specific column names. The mapping from
`distsamp` output is mechanical:

```{r, eval = FALSE}
library(dsm)

segment.data <- data.frame(
  Sample.Label = segs$segments$seg_id,
  Effort       = segs$segments$seg_eff,   # km
  X            = segs$segments$mid_lon,   # project these first, see below
  Y            = segs$segments$mid_lat,
  beaufort     = segs$segments$wt_beaufort
)

observation.data <- data.frame(
  object       = seq_len(nrow(df_data)),
  Sample.Label = df_data$seg_id,
  size         = df_data$size,
  distance     = df_data$distance
)
```

`object` is just a detection identifier; `dsm` uses it to link a detection to
its row in the fitted detection function, so it must match the row order of the
data you passed to `ds()`.

### Project the coordinates first

`mid_lon`/`mid_lat` are decimal degrees. Fitting a smooth of longitude and
latitude in degrees distorts distances badly at the latitudes NARWC surveys
cover — a degree of longitude is about 81 km at 43°N against 111 km for a degree
of latitude. Project to something equal-area before modelling:

```{r, eval = requireNamespace("sf", quietly = TRUE)}
pts <- segments_as_sf(segs, "midpoints")

# Albers equal-area, Gulf of Maine
gom_aea <- paste(
  "+proj=aea +lat_1=40 +lat_2=46 +lat_0=43 +lon_0=-69",
  "+x_0=0 +y_0=0 +datum=WGS84 +units=km +no_defs"
)
projected <- sf::st_transform(pts, gom_aea)

xy <- sf::st_coordinates(projected)
head(xy)
```

Then use those `X`/`Y` in `segment.data`.

## Fitting the model

```{r, eval = FALSE}
mod <- dsm(
  count ~ s(X, Y, bs = "ts") + s(depth, bs = "ts"),
  ddf.obj = df,
  segment.data = segment.data,
  observation.data = observation.data,
  family = tw(),          # Tweedie handles the zero inflation
  method = "REML"
)

summary(mod)
plot(mod)
gam.check(mod)
```

A Tweedie or negative binomial family is usually right: most segments hold zero
animals, and the non-zero ones are overdispersed. Miller et al. (2013) discuss
the choice.

## Covariates are your job

`distsamp` gives you a location per segment and nothing else. Environmental
covariates — depth, sea surface temperature, fronts — have to be sampled at each
segment midpoint yourself. `segments_as_sf(segs, "midpoints")` is the intended
handoff:

```{r, eval = FALSE}
pts <- segments_as_sf(segs, "midpoints")

# e.g. with terra
depth <- terra::extract(bathymetry_raster, terra::vect(pts))
segment.data$depth <- depth[, 2]
```

Two cautions.

Sample at the **midpoint**, which is what Becker et al. (2019) do — covariates
"derived based on the segment's geographical mid-point", with SST and depth
standard deviations taken over a 3 × 3-pixel box around it. `segment_midpoints()`
places that point half the segment's *effort* along the track rather than at the
centroid of its records, so it is on the trackline by construction.

And match covariates in **time** as well as space. A segment flown in April
needs April's SST, not the annual mean. `segs$segments$DATE` carries the survey
date for exactly this.

## Reporting

Record the `seed`. Without it the segmentation is not reproducible, and neither
is anything downstream of it.

```{r}
str(segs$settings)
```

Suggested methods wording is in `docs/06-references.md` in the package
repository.

## References

Becker, E.A., Forney, K.A., Ferguson, M.C., Foley, D.G., Smith, R.C., Barlow, J.
and Redfern, J.V. (2010) Comparing California Current cetacean-habitat models
developed using in situ and remotely sensed sea surface temperature data.
*Marine Ecology Progress Series* 413:163-183.
<https://doi.org/10.3354/meps08696>

Becker, E.A., Forney, K.A., Redfern, J.V., Barlow, J., Jacox, M.G., Roberts, J.J.
and Palacios, D.M. (2019) Predicting cetacean abundance and distribution in a
changing climate. *Diversity and Distributions* 25:626-643.
<https://doi.org/10.1111/ddi.12867>

Hedley, S.L. and Buckland, S.T. (2004) Spatial models for line transect sampling.
*Journal of Agricultural, Biological, and Environmental Statistics* 9:181-199.
<https://doi.org/10.1198/1085711043578>

Miller, D.L., Burt, M.L., Rexstad, E.A. and Thomas, L. (2013) Spatial models for
distance sampling data: recent developments and future directions. *Methods in
Ecology and Evolution* 4:1001-1010.
<https://doi.org/10.1111/2041-210X.12105>

Miller, D.L., Rexstad, E., Thomas, L., Marshall, L. and Laake, J.L. (2019)
Distance sampling in R. *Journal of Statistical Software* 89(1):1-28.
<https://doi.org/10.18637/jss.v089.i01>
