---
title: "Segmenting NARWC line-transect data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Segmenting NARWC line-transect data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(distsamp)
```

## What this package does

Density surface modelling needs survey effort chopped into short, roughly equal
pieces — segments — each with a location, an amount of effort, and a count of
what was seen along it. This package takes aerial line-transect survey data in
the format of the North Atlantic Right Whale Consortium (NARWC) sightings
database and produces exactly that.

The input contract is the NARWC format described in Kenney (2023), *The North
Atlantic Right Whale Consortium Database: A Guide for Users and Contributors*,
Version 8. The output is a segment table you can hand to `Distance` or `dsm`.

## Reading data

`read_narwc()` standardises column names and types. It accepts the handbook's
canonical `LAT_DD`/`LONG_DD` as well as the `LATITUDE`/`LONGITUDE` spelling used
in the extracts the database manager distributes.

```{r}
path <- system.file("extdata", "narwc-example.csv", package = "distsamp")
dat <- read_narwc(path)

dim(dat)
head(dat[, c("FILEID", "EVENTNO", "TIME", "LATITUDE", "LEGTYPE", "LEGSTAGE", "SPECCODE")])
```

**Everything in this vignette runs on mock data.** `narwc-example.csv` is a
synthetic two-day survey generated by `data-raw/make-fixture.R`, built to mirror
the hypothetical survey in Figure 2 of the handbook: transits, survey lines,
sightings sharing an event number, a pilot sighting, a cross-leg, a break-off to
circle, and a line abandoned and re-flown later in the day. The positions,
dates, and animals are invented — the geometry is deliberately simple, with
every line running due north along a constant meridian in 0.01-degree steps, so
that expected distances can be checked by hand.

No real NARWC records are distributed with this package, and none have been used
to develop it. Treat every figure here as illustrating the mechanics, never as
a result. Real data must be requested from the
[NARWC](https://www.narwc.org/sightings-database.html), and validating this
package against a real extract is still an open item.

## Data that is not quite handbook shape

Two things real extracts do that the bundled fixture does not.

### Columns from a particular survey programme

Survey programmes add their own derived columns. `read_narwc()` keeps the
handbook columns and drops the rest — but it tells you what it dropped, rather
than discarding it silently:

```{r}
ccs <- read.csv(path)
ccs$Tr_SIGHTING <- 1
ccs$IS_LAT <- ccs$LAT_DD
ccs$IS_LONG <- ccs$LONG_DD

kept <- read_narwc(ccs)
```

`narwc_profiles()` records what is known about those columns, per programme,
including how confident that knowledge is:

```{r}
narwc_profiles("ccs")[, c("column", "role", "confidence")]
```

Naming the profile keeps them:

```{r}
names(read_narwc(ccs, profile = "ccs"))[10:14]
```

Keeping is not interpreting. A profile is never applied because a column name
matched, and `distsamp` does nothing with these columns beyond carrying them:
`Tr_SIGHTING` means "sighting made from the track-line" in a CCS file, and there
is nothing stopping another programme using that name for something else. A
column name is not a contract between programmes.

### Blank rows where a value has not changed

NARWC often records a value once and leaves it blank until it changes — `LEGTYPE`
is entered at the start of a census line and the rows beneath are empty until the
leg type changes. `fill_narwc()` carries that state forward:

```{r}
gappy <- data.frame(
  FILEID = "A", DATE = as.Date("2024-04-01"), EVENTNO = 1:6,
  LEGTYPE  = c(2, NA, NA, NA, 1, NA),
  BEAUFORT = c(2, NA, NA, 4, NA, NA)
)
fill_narwc(gappy)
```

Three things it does that a bare `tidyr::fill()` does not.

**It groups by file and day.** Ungrouped, a sea state from the last record of
one survey day carries into the next, and a leg number crosses into a different
survey entirely. A filled value is indistinguishable from a recorded one
afterwards, so this is not a mistake you can find later.

**It refuses to fill sighting columns.** Carrying `SPECCODE` and `NUMBER` forward
would replicate one group of three right whales onto every row until the next
sighting:

```{r, error = TRUE}
fill_narwc(gappy, columns = "NUMBER")
```

**It separates recovery from inference.** `"downup"` fills down and then up, so
its backward fills are the values before the first record of a group — state
that was never logged. Those are counted separately in the report, because only
the forward fills recover something that was there.

One caution. Filling `LEGSTAGE` is the least safe of the defaults: if a file
records `1` (begin line) and leaves the continuation rows blank, filling down
marks the whole line "begin line" rather than `2` (continue) — and since
on-effort eligibility is `LEGSTAGE == 2`, that line drops out of every distance
calculation. Check the per-column counts against what you expect.

## Checking data

`validate_narwc()` reports problems rather than raising them, so you can decide
what matters. It never modifies the data.

```{r}
validate_narwc(dat)
```

Nothing to report here. On real data you will typically see codes outside the
handbook's code book, sightings recorded at line-boundary events, and gaps in
required fields. Each finding names the rows involved:

```{r}
broken <- dat
broken$LEGTYPE[1] <- 8       # not a NARWC LEGTYPE
broken$LONGITUDE <- abs(broken$LONGITUDE)  # sign convention lost

issues <- validate_narwc(broken)
issues[, c("check", "severity", "column", "n")]
issues$message[1]
```

## Segmenting

`segment_survey()` runs the whole pipeline.

```{r}
segs <- segment_survey(dat, seg_length = 5, seed = 1)
segs
```

The `segments` table has one row per segment:

```{r}
head(segs$segments[, c("seg_id", "LEGNO3", "seg_eff", "mid_lat", "mid_lon", "wt_beaufort")])
```

`seg_eff` is the segment's realised length in km. `mid_lat`/`mid_lon` are the
point half of the segment's effort along the track — the location to use for a
covariate lookup. `wt_beaufort` is the mean sea state weighted by the distance
each record covers, which is the right form for a detection covariate.

Counts arrive separately, one row per segment per species:

```{r}
head(segs$sightings)
```

For a density surface model you usually want them spread across columns:

```{r}
wide <- segments_wide(segs)
head(wide[, c("seg_id", "seg_eff", "mid_lat", "mid_lon", "n_RIWH", "n_FIWH")])
```

## Effort is conserved

Every kilometre of on-effort trackline ends up in exactly one segment.

```{r}
sum(segs$tracks$track_effort)
sum(segs$segments$seg_eff)
```

## Reproducibility

Segmentation makes two random choices: which segment absorbs a track's leftover
distance, and whether a segment that cannot land exactly on its target runs a
little long or a little short. Pass a `seed` and a run repeats exactly.

```{r}
a <- segment_survey(dat, seg_length = 5, seed = 1)
b <- segment_survey(dat, seg_length = 5, seed = 1)
identical(a$segments, b$segments)

c <- segment_survey(dat, seg_length = 5, seed = 2)
identical(a$segments$seg_eff, c$segments$seg_eff)
```

Your session's own random stream is never disturbed.

## Choosing a distance method

```{r}
dist_methods()
```

Becker's `segchopr` routine and Kenney and Winn (1986) are the same formula —
the spherical law of cosines, scaled by 111.12 km per degree — so choosing
between them does not change your results:

```{r}
gc_distance(43, -69, 44, -70, method = "becker") ==
  gc_distance(43, -69, 44, -70, method = "kenney")
```

Both names are kept so that existing scripts read sensibly and so you can record
which lineage you meant to follow. The default is `"haversine"`, which is the
same sphere but numerically stable at the short ranges between consecutive
survey records, where the law of cosines loses precision.

```{r}
segment_survey(dat, seg_length = 5, seed = 1, dist_method = "becker")$settings$dist_method
```

## What counts as effort

`flag_effort()` decides which records are on effort. The defaults follow the
CETAP standard: on a survey line (`LEGTYPE == 2`), Beaufort at most 3, below
1,200 ft, and visibility at least 2 nautical miles.

```{r}
table(flag_effort(dat)$OnOff.Effort)
table(flag_effort(dat, max_beaufort = 1)$OnOff.Effort)
```

Visibility deserves a note. `VISIBLTY` carries two encodings in one column: since
2004 an actual distance in nautical miles, and before that a code folded into the
same field as a negative number. `-1` means *clear for at least 2 nautical
miles* — good conditions — so a plain `VISIBLTY >= 2` test throws away every
legacy record.

```{r}
visibility_ok(c(5, 1.5, -1, -2))
```

## What counts as a sighting

Only detections that arose from standard search effort belong in a density
estimate. By default the package excludes sightings by anyone other than an
on-duty observer (`LEGSTAGE == 6`), sightings found afterwards in a vertical
photograph (`LEGSTAGE == 7`), and identifications that are merely possible or
unrecorded (`IDREL` of 1 or 9).

```{r}
strict <- segment_survey(dat, seg_length = 5, seed = 1)
loose <- segment_survey(
  dat, seg_length = 5, seed = 1,
  sighting_args = list(legstage_exclude = numeric(0), idrel_keep = 1:3)
)

sum(strict$sightings$n_sightings)
sum(loose$sightings$n_sightings)
```

A group sighted from the track and then circled for photographs has its position
recorded off effort. Those detections are attached back to the segment they came
from, following the CETAP rule that only further groups *of the same species*
count with the original:

```{r}
sapply(c("none", "same_species", "all"), function(m) {
  s <- segment_survey(dat, seg_length = 5, seed = 1, circling = m)
  sum(s$sightings$n_animals[s$sightings$SPECCODE == "RIWH"])
})
```

## Perpendicular distances

Surveys that record the `ANGLEL`/`ANGLER` declination angles get perpendicular
distances computed automatically, and a `detections` table in the shape a
detection function wants.

```{r}
head(segs$detections)
```

The geometry is handbook 8.A.2: the angle is measured below the horizon at the
moment the sighting is abeam, so the distance is `ALT / tan(angle)`.

```{r}
# 45 degrees below the horizon at 229 m puts the animal 229 m off the track
perp_distance(45, altitude = 229)
perp_distance(c(75, 60, 45, 30, 15), altitude = 229)
```

These replaced `STRIP` in 2022: survey altitudes had to rise once offshore wind
turbines were in place, and `STRIP`'s fixed distance intervals shift with
altitude while an angle does not.

### The other three sources

Angles only exist from 2022. The archive records right-angle distance three other
ways, and each has its own function.

**`STRIP`, before 2022.** An *interval*, not a point, read off calibrated
markings on the observation bubble. Handbook 8.A.31 defines two code books, and
which applies depends on the programme, the date, and the aircraft — so the same
code means different things:

```{r}
strip_distance(c(5, 13), scheme = "cetap", units = "nmi")
strip_distance(c(5, 13), scheme = "nlpsc", units = "nmi")
```

Code 13 is 1–2 nmi under CETAP and over 4 nmi under NLPSC. Reading a code with
the wrong book produces plausible distances that are wrong, and nothing
downstream would notice — so with `scheme = "auto"` the book is chosen from the
survey date, and with neither a date nor a scheme it errors rather than guess.
`narwc_strip_bins()` prints either book in full.

**Exact positions, where `S_LAT`/`S_LONG` exist.** The perpendicular distance can
then be measured rather than inferred. `exact_distance()` projects the sighting
onto the track-line:

```{r}
prepped <- flag_effort(make_leg_id(dat))
ex <- exact_distance(prepped)
subset(cbind(prepped["SPECCODE"], ex), !is.na(distance))
```

`along` is the point of that table. It is how far the aircraft was from being
abeam when the sighting was logged. Measuring straight to the animal instead —
which is what a naive implementation does — gives
`sqrt(distance^2 + along^2)`, always too large and never too small, which widens
the fitted strip and biases density down. The row with `along` of 300 m has a
radial distance more than twice its perpendicular one.

**Sightings logged while circling.** These are usually real on-effort detections:
the animal was seen from the track-line, which is why the aircraft left it.
`circling_distance()` ties each back to the `LEGSTAGE == 3` break-off record and
measures from there:

```{r}
circ <- circling_distance(flag_circling(make_leg_id(dat)))
subset(circ, !is.na(distance))
```

`radial` is the straight-line distance from the break-off point; `distance` is
that projected onto the census line. Treat these as a weaker source than the
other three — the position is fixed minutes after detection, by which time the
animal has moved — and exclude them from a detection function unless including
them is a deliberate, stated choice.

Two things to watch:

**Units.** `ALT` is in metres, so distances are metres by default, but `seg_eff`
is in kilometres. Passing both to `Distance::ds()` without reconciling them
scales density by 1000. Use `distance_units = "km"`, or supply `convert_units`.

```{r}
segment_survey(dat, seg_length = 5, seed = 1,
               distance_units = "km")$detections$distance[1:3]
```

**Circling detections have no distance.** They still count towards a segment's
abundance, but a position logged off the track is not a perpendicular distance
and must not be fitted:

```{r}
subset(segs$detections, circling == 1)
```

Distances are restricted to on-effort census records by default, because
handbook 8.A.31 notes that some teams record angles during transits and circling
for practice — those must not enter a detection function.

## Looking at the result

Four diagnostic views, each answering something the tables make hard to see.
They need `ggplot2`, and return ordinary `ggplot` objects.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
plot(segs)
```

The track flown, the segment midpoints sized by effort, and the sightings. A
segmentation that has gone wrong is usually obvious here — midpoints strung
along a line the aircraft never flew, or bunched where a track should have split.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
plot(segs, what = "tracks")
```

The same positions coloured by `new_trackno`. This is the view that shows
whether track splitting worked: a break in effort should start a new colour, and
a colour should never span two places the aircraft could not have flown between.
The fixture's re-flown line is the case to look at.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
plot(segs, what = "effort")
```

Segment lengths against the target, with the tolerance band dashed. A spread is
expected — the Becker method gives each track one absorbing segment that takes
the remainder — but mass outside the band is not.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
plot(segs, what = "distances")
```

The detection-function diagnostic, worth looking at before `Distance::ds()` is
called. Look for a shoulder near zero and a tail that falls away.

Two cautions on reading it. A dip in the first bin need not be a sampling
artefact: an aircraft cannot see the water directly beneath it, and for the
Skymaster the handbook says so outright — a reason to truncate on the left, not
to assume the animals were absent. And a smooth curve is not evidence that
`g(0) = 1`; animals submerged when the aircraft passed leave no trace in this
plot at all.

For a real map with coastlines, hand the positions to `sf` instead.

## Spatial output

```{r, eval = requireNamespace("sf", quietly = TRUE)}
pts <- segments_as_sf(segs, "midpoints")
lines <- segments_as_sf(segs, "lines")
lines[1:3, c("seg_id", "seg_eff")]
```

Cropping works on the whole result at once, keeping the tables consistent:

```{r}
gom <- c(xmin = -69.05, xmax = -68.5, ymin = 42, ymax = 48)
nrow(crop_to_bbox(segs, gom)$segments)
```

## Writing results

Nothing in the package writes to disk unless you ask.

```{r}
out <- file.path(tempdir(), "segments")
write_segments(segs, out)
list.files(out)
```

## Running the steps yourself

Every stage is exported, so you can intervene between them:

```{r}
step <- dat |>
  make_leg_id() |>          # separate re-occupations of a line
  flag_circling() |>        # mark records taken while circling
  flag_effort() |>          # decide what is on effort
  point_to_point_effort() |># great-circle distance along the track
  split_tracks()            # start a new track where effort breaks

tracks <- track_effort(step)
plan <- plan_segments(tracks, seg_length = 5, seed = 1)
head(plan)
```

## Citing this

The segmentation method is Becker et al. (2010). The package is the
implementation. Cite them separately, and record the `seed` — without it the
segmentation is not reproducible.

Note that the published methods state the target segment length and that segments
are cut from continuous effort, but do not describe what happens to the leftover
distance at the end of a track. The tolerance test and the random choice of an
absorbing segment come from Becker's `segchopr` code; `?plan_segments` is the
documentation of record for them.

## References

Becker, E.A., Forney, K.A., Ferguson, M.C., Foley, D.G., Smith, R.C., Barlow, J.
and Redfern, J.V. (2010) Comparing California Current cetacean-habitat models
developed using in situ and remotely sensed sea surface temperature data. *Marine
Ecology Progress Series* 413:163-183. <https://doi.org/10.3354/meps08696>

Becker, E.A., Forney, K.A., Redfern, J.V., Barlow, J., Jacox, M.G., Roberts, J.J.
and Palacios, D.M. (2019) Predicting cetacean abundance and distribution in a
changing climate. *Diversity and Distributions* 25:626-643.
<https://doi.org/10.1111/ddi.12867>

Hedley, S.L. and Buckland, S.T. (2004) Spatial models for line transect sampling.
*Journal of Agricultural, Biological, and Environmental Statistics* 9:181-199.
<https://doi.org/10.1198/1085711043578>

Miller, D.L., Burt, M.L., Rexstad, E.A. and Thomas, L. (2013) Spatial models for
distance sampling data: recent developments and future directions. *Methods in
Ecology and Evolution* 4:1001-1010.
<https://doi.org/10.1111/2041-210X.12105>

Kenney, R.D. (2023) *The North Atlantic Right Whale Consortium Database: A Guide
for Users and Contributors, Version 8.* NARWC Reference Document 2023-01.

Kenney, R.D. and Winn, H.E. (1986) Cetacean high-use habitats of the northeast
United States continental shelf. *Fishery Bulletin* 84:345-357.
