---
title: "Segmenting NARWC line-transect data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Segmenting NARWC line-transect data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(distsamp)
```

## What this package does

Density surface modelling needs survey effort chopped into short, roughly equal
pieces — segments — each with a location, an amount of effort, and a count of
what was seen along it. This package takes aerial line-transect survey data in
the format of the North Atlantic Right Whale Consortium (NARWC) sightings
database and produces exactly that.

The input contract is the NARWC format described in Kenney (2021), *The North
Atlantic Right Whale Consortium Database: A Guide for Users and Contributors*,
Version 7. The output is a segment table you can hand to `Distance` or `dsm`.

## Reading data

`read_narwc()` standardises column names and types. It accepts the handbook's
canonical `LAT_DD`/`LONG_DD` as well as the `LATITUDE`/`LONGITUDE` spelling used
in the extracts the database manager distributes.

```{r}
path <- system.file("extdata", "narwc-example.csv", package = "distsamp")
dat <- read_narwc(path)

dim(dat)
head(dat[, c("FILEID", "EVENTNO", "TIME", "LATITUDE", "LEGTYPE", "LEGSTAGE", "SPECCODE")])
```

The bundled example is a synthetic two-day survey built to mirror the
hypothetical survey in Figure 2 of the handbook: transits, survey lines,
sightings sharing an event number, a pilot sighting, a cross-leg, a break-off to
circle, and a line abandoned and re-flown later in the day.

## Checking data

`validate_narwc()` reports problems rather than raising them, so you can decide
what matters. It never modifies the data.

```{r}
validate_narwc(dat)
```

Nothing to report here. On real data you will typically see codes outside the
handbook's code book, sightings recorded at line-boundary events, and gaps in
required fields. Each finding names the rows involved:

```{r}
broken <- dat
broken$LEGTYPE[1] <- 8       # not a NARWC LEGTYPE
broken$LONGITUDE <- abs(broken$LONGITUDE)  # sign convention lost

issues <- validate_narwc(broken)
issues[, c("check", "severity", "column", "n")]
issues$message[1]
```

## Segmenting

`segment_survey()` runs the whole pipeline.

```{r}
segs <- segment_survey(dat, seg_length = 5, seed = 1)
segs
```

The `segments` table has one row per segment:

```{r}
head(segs$segments[, c("seg_id", "LEGNO3", "seg_eff", "mid_lat", "mid_lon", "wt_beaufort")])
```

`seg_eff` is the segment's realised length in km. `mid_lat`/`mid_lon` are the
point half of the segment's effort along the track — the location to use for a
covariate lookup. `wt_beaufort` is the mean sea state weighted by the distance
each record covers, which is the right form for a detection covariate.

Counts arrive separately, one row per segment per species:

```{r}
head(segs$sightings)
```

For a density surface model you usually want them spread across columns:

```{r}
wide <- segments_wide(segs)
head(wide[, c("seg_id", "seg_eff", "mid_lat", "mid_lon", "n_RIWH", "n_FIWH")])
```

## Effort is conserved

Every kilometre of on-effort trackline ends up in exactly one segment.

```{r}
sum(segs$tracks$track_effort)
sum(segs$segments$seg_eff)
```

## Reproducibility

Segmentation makes two random choices: which segment absorbs a track's leftover
distance, and whether a segment that cannot land exactly on its target runs a
little long or a little short. Pass a `seed` and a run repeats exactly.

```{r}
a <- segment_survey(dat, seg_length = 5, seed = 1)
b <- segment_survey(dat, seg_length = 5, seed = 1)
identical(a$segments, b$segments)

c <- segment_survey(dat, seg_length = 5, seed = 2)
identical(a$segments$seg_eff, c$segments$seg_eff)
```

Your session's own random stream is never disturbed.

## Choosing a distance method

```{r}
dist_methods()
```

Becker's `segchopr` routine and Kenney and Winn (1986) are the same formula —
the spherical law of cosines, scaled by 111.12 km per degree — so choosing
between them does not change your results:

```{r}
gc_distance(43, -69, 44, -70, method = "becker") ==
  gc_distance(43, -69, 44, -70, method = "kenney")
```

Both names are kept so that existing scripts read sensibly and so you can record
which lineage you meant to follow. The default is `"haversine"`, which is the
same sphere but numerically stable at the short ranges between consecutive
survey records, where the law of cosines loses precision.

```{r}
segment_survey(dat, seg_length = 5, seed = 1, dist_method = "becker")$settings$dist_method
```

## What counts as effort

`flag_effort()` decides which records are on effort. The defaults follow the
CETAP standard: on a survey line (`LEGTYPE == 2`), Beaufort at most 3, below
1,200 ft, and visibility at least 2 nautical miles.

```{r}
table(flag_effort(dat)$OnOff.Effort)
table(flag_effort(dat, max_beaufort = 1)$OnOff.Effort)
```

Visibility deserves a note. `VISIBLTY` carries two encodings in one column: since
2004 an actual distance in nautical miles, and before that a code folded into the
same field as a negative number. `-1` means *clear for at least 2 nautical
miles* — good conditions — so a plain `VISIBLTY >= 2` test throws away every
legacy record.

```{r}
visibility_ok(c(5, 1.5, -1, -2))
```

## What counts as a sighting

Only detections that arose from standard search effort belong in a density
estimate. By default the package excludes sightings by anyone other than an
on-duty observer (`LEGSTAGE == 6`), sightings found afterwards in a vertical
photograph (`LEGSTAGE == 7`), and identifications that are merely possible or
unrecorded (`IDREL` of 1 or 9).

```{r}
strict <- segment_survey(dat, seg_length = 5, seed = 1)
loose <- segment_survey(
  dat, seg_length = 5, seed = 1,
  sighting_args = list(legstage_exclude = numeric(0), idrel_keep = 1:3)
)

sum(strict$sightings$n_sightings)
sum(loose$sightings$n_sightings)
```

A group sighted from the track and then circled for photographs has its position
recorded off effort. Those detections are attached back to the segment they came
from, following the CETAP rule that only further groups *of the same species*
count with the original:

```{r}
sapply(c("none", "same_species", "all"), function(m) {
  s <- segment_survey(dat, seg_length = 5, seed = 1, circling = m)
  sum(s$sightings$n_animals[s$sightings$SPECCODE == "RIWH"])
})
```

## Spatial output

```{r, eval = requireNamespace("sf", quietly = TRUE)}
pts <- segments_as_sf(segs, "midpoints")
lines <- segments_as_sf(segs, "lines")
lines[1:3, c("seg_id", "seg_eff")]
```

Cropping works on the whole result at once, keeping the tables consistent:

```{r}
gom <- c(xmin = -69.05, xmax = -68.5, ymin = 42, ymax = 48)
nrow(crop_to_bbox(segs, gom)$segments)
```

## Writing results

Nothing in the package writes to disk unless you ask.

```{r}
out <- file.path(tempdir(), "segments")
write_segments(segs, out)
list.files(out)
```

## Running the steps yourself

Every stage is exported, so you can intervene between them:

```{r}
step <- dat |>
  make_leg_id() |>          # separate re-occupations of a line
  flag_circling() |>        # mark records taken while circling
  flag_effort() |>          # decide what is on effort
  point_to_point_effort() |># great-circle distance along the track
  split_tracks()            # start a new track where effort breaks

tracks <- track_effort(step)
plan <- plan_segments(tracks, seg_length = 5, seed = 1)
head(plan)
```

## Citing this

The segmentation method is Becker et al. (2010). The package is the
implementation. Cite them separately, and record the `seed` — without it the
segmentation is not reproducible.

Note that the published methods state the target segment length and that segments
are cut from continuous effort, but do not describe what happens to the leftover
distance at the end of a track. The tolerance test and the random choice of an
absorbing segment come from Becker's `segchopr` code; `?plan_segments` is the
documentation of record for them.

## References

Becker, E.A., Forney, K.A., Ferguson, M.C., Foley, D.G., Smith, R.C., Barlow, J.
and Redfern, J.V. (2010) Comparing California Current cetacean-habitat models
developed using in situ and remotely sensed sea surface temperature data. *Marine
Ecology Progress Series* 413:163-183. <https://doi.org/10.3354/meps08696>

Becker, E.A., Forney, K.A., Redfern, J.V., Barlow, J., Jacox, M.G., Roberts, J.J.
and Palacios, D.M. (2019) Predicting cetacean abundance and distribution in a
changing climate. *Diversity and Distributions* 25:626-643.
<https://doi.org/10.1111/ddi.12867>

Hedley, S.L. and Buckland, S.T. (2004) Spatial models for line transect sampling.
*Journal of Agricultural, Biological, and Environmental Statistics* 9:181-199.
<https://doi.org/10.1198/1085711043578>

Miller, D.L., Burt, M.L., Rexstad, E.A. and Thomas, L. (2013) Spatial models for
distance sampling data: recent developments and future directions. *Methods in
Ecology and Evolution* 4:1001-1010.
<https://doi.org/10.1111/2041-210X.12105>

Kenney, R.D. (2021) *The North Atlantic Right Whale Consortium Database: A Guide
for Users and Contributors, Version 7.* NARWC Reference Document 2021-01.

Kenney, R.D. and Winn, H.E. (1986) Cetacean high-use habitats of the northeast
United States continental shelf. *Fishery Bulletin* 84:345-357.
